Appendix to "Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model

### 2023-04-03 src/docs_paper/appendix.Rmd
Add multinomial to main text, many changes to appendix including table of number of latent field and hyperparameters, and table connecting TMB notation to mathematical notation

<!-- So that the tables and figures in this appendix are prefixed by "A." -->
\setcounter{table}{0}  
\renewcommand{\thetable}{A.\arabic{table}} 
\setcounter{figure}{0} 
\renewcommand{\thefigure}{A.\arabic{figure}}

# Simplified Naomi model description

In this section we describes more fully the simplified Naomi model considered in the main text.
To do so, we draw substantially from the supplementary material of @eaton2021naomi.

@@

Let $\nu$ the most recent national household survey with HIV testing which has taken place in the country of interest at time $t = T_1$
Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age bands.
As short-hand, we write $a = l$ to refer to the age band with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
We index the following quantities by district, sex and age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$, annual HIV incidence rate $\lambda_{x, s, a}$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a}$.
Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.

@@

Let $u$ be a generic random effect.
We use use structured random effects to share information across units assessed as being similar, such as neighbouring districts or adjacent age-bands.
We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
$$
u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
$$
where $u_{-i}$ is $u$ with the $i$th unit removed, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v^\star$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w^\star$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
$$
u = \sigma \left( \sqrt\phi \cdot v^\star + \sqrt{1 - \phi} \cdot w^\star \right).
$$
We specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align*}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_t, \quad i = 2, \ldots
\end{align*}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, $|\rho| < 1$ is the lag-one correlation parameter.

@@

We assume the household survey was run according to a complex survey design where each individual $j \in U$ has probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$.
Suppose we observe an outcome $y_j$ for $j \in S$.
Let $w_j = 1 / \pi_j$ be design weights, then the weighted mean
$$
\hat y = \frac{\sum_{j \in S} w_j y_j}{\sum_{j \in S} w_j}
$$
Define Kish ESS.
We make these computations using the `survey` \textsc{R} package [@lumley2004analysis].

@@

@@

We model HIV prevalence on the logit scale using the linear predictor
$$
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
$$
where $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
We use the prior distributions $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.

@@

We model ART coverage using the linear predictor
$$
\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + u^\alpha_a + u_a^{\alpha, s = \text{M}} + u^\alpha_x + u_x^{\alpha, s = \text{M}} + u_x^{\alpha, a < 15} + \eta^\alpha_{R_x, s, a} 
$$
with terms and priors analogous to the HIV prevalence process model in Section \ref{sec:hiv-prev} above.

### HIV incidence rate

We model HIV incidence rate using the linear predictor
$$
\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda, \label{eq:hiv-inc}
$$
where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women, $\rho_{x}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences
$$
\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}, \label{eq:rho-agg}
$$
$\alpha_{x}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages
$$
\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}, \label{eq:alpha-agg}
$$
$\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation], $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output.
Note that the HIV incidence rate process (Equation \ref{eq:hiv-inc}) depends on those of the HIV prevalence (Section \ref{sec:hiv-prev}) and the ART coverage (Section \ref{sec:art-cov}) via the adult aggregate indicators (Equations \ref{eq:rho-agg} and \ref{eq:alpha-agg}) violating the conditions required being within the latent Gaussian model (LGM) class as defined by @rue2009approximate.

### ANC testing

HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women are modelled as being offset on the logit-scale from the corresponding district-age indicators $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ according to
\begin{align}
\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \label{eq:anc-hiv-prev} \\
\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}} \label{eq:anc-art-cov},
\end{align}
where, for $\theta \in \{\rho, \alpha\}$, $\beta^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, 5)$ are the average differences between population and ANC outcomes, $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ are IID district random effects with $\sigma_X^{\theta^{\text{ANC}}} \sim \mathcal{N}^+(0, 1)$, and $\eta_{R_x, a}^{\theta^{\text{ANC}}}$ for are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$.
Dependence here too on $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ in Equations \ref{eq:anc-hiv-prev} and \ref{eq:anc-art-cov} violates the conditions required to be a LGM.

### ART attendance

@@

Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ recieves ART in district $x'$.
We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district or its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation, and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.
To model $\gamma_{x, x'}$ for $x \sim x'$ we use a multinomial logistic regression model, based on the log-odds ratios
\begin{equation}
\tilde \gamma_{x, x'} = \log \left( \frac{\gamma_{x, x'}}{1 - \gamma_{x, x'}} \right) = \tilde \gamma_0 + u_x^{\tilde \gamma}, \label{eq:log-or}
\end{equation}
where $\tilde \gamma_0 = -4$ is a fixed intercept, and $u_x^{\tilde \gamma} \sim \mathcal{N}(0, \sigma_X^{\tilde \gamma})$ are district random effects with $\sigma_X^{\tilde \gamma} \sim \mathcal{N}^+(0, 2.5)$.
Note that Equation \ref{eq:log-or} does not depend on $x'$, such that $\gamma_{x, x'}$ is only a function of $x$.
Choice of $\tilde \gamma_0 = -4$ implies a prior mean on $\gamma_{x, x'}$ of 1.8%, such that $(100 - 1.8 \times \text{ne}(x))\%$ of ART clients in district $x$ obtain treatment in their home district, a-priori.
We fix $\tilde \gamma_{x, x} = 0$ and recover the multinomial probabilities using the softmax
$$
\gamma_{x, x'} = \frac{\exp(\tilde \gamma_{x, x'})}{\sum_{x^\star \in \{x, \text{ne}(x)\}} \exp(\tilde \gamma_{x, x^\star})}.
$$
Given the total number of PLHIV on ART $A_{x, s, a} = N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}$, the number of ART clients who reside in district $x$ and obtain ART in district $x'$ are $A_{x, x', s, a} = A_{x, s, a} \cdot \gamma_{x, x'}$, and the total attending ART facilities in district $x'$ are
$$
\tilde A_{x', s, a} = \sum_{x \in \{x', \text{ne}(x')\}} A_{x, x', s, a}.
$$

## Likelihood specification

@@

For $\theta \in \{\rho, \alpha, \kappa\}$, the household survey $\nu$ furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}}$, and observed number of cases
$$
y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}}.
$$

For HIV prevalence and ART coverage we use the binomial working likelihoods
\begin{align}
y^{\hat \rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \\
\rho_{\{x\}, \{s\}, \{a\}} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a}},
\end{align}
and
\begin{align}
y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \\
\alpha_{\{x\}, \{s\}, \{a\}} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
\end{align}

>>>>

```{=tex}
\setcounter{table}{0}  
\renewcommand{\thetable}{S\arabic{table}} 
\setcounter{figure}{0} 
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{section}{0} 
\renewcommand{\thesection}{S\arabic{section}}
```

# Simplified Naomi model description \label{sec:math}

In this section we describe more fully the simplified Naomi model considered in the main text. In doing so we draw from the supplementary material of @eaton2021naomi.

@@

Let $\nu$ the most recent national household survey with HIV testing which has taken place in the country of interest.
Let $x \in \{1, \ldots, n\}$ refer to a district located within the Spectrum [@stover2010spectrum] region $R_x$.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, \text{80+}\}$ be five-year age bands.
As short-hand, we write $a = l$ to refer to the age band with lower bound $l$, e.g. $a = 20$ for $a = \text{20-25}$.
We index the known quantity population size $N_{x, s, a}$ by district, sex and age-band, as well as the following unknown quantities: HIV prevalence $\rho_{x, s, a} \in [0, 1]$, ART coverage $\alpha_{x, s, a} \in [0, 1]$, annual HIV incidence rate $\lambda_{x, s, a} > 0$, and proportion of HIV positive persons recently infected $\kappa_{x, s, a} \in [0, 1]$.
Sometimes data are observed at an aggregate level, rather than the more granular modelled level.
In this instance, we use $\{\cdot\}$ to generically refer to a aggregate set over which an observation is made, e.g. $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for the adult age range 15-49.
In the main text we use a more concise, but less descriptive, approach.
Finally, we let $\sum_{\{x\}}$ be used as shorthand for $\sum_{x \in \{x\}}$, and likewise for $s$ and $a$.

@@

Let $u$ be a generic random effect. We use use structured random effects to enable partial pooling of information across units assessed as being similar, such as neighbouring districts or adjacent age-bands. We specify the first order auto-regressive model by $u \sim \text{AR}1(\sigma, \phi)$ such that
\begin{align*}
u_1 &\sim \left( 0, \frac{1}{1 - \rho^2} \right), \\
u_i &= \rho u_{i - 1} + \epsilon_t, \quad i = 2, \ldots
\end{align*}
where $\epsilon_i \sim \mathcal{N}(0, 1)$ is Gaussian white noise, and $|\rho| < 1$ is the lag-one correlation parameter.
We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with full conditionals
$$
u_i \, | \, u_{-i} \sim \mathcal{N} \left(\frac{\sum_{j: j \sim i} u_j}{n_{\delta i}}, \frac{\sigma^2}{n_{\delta i}}\right),
$$
where $u_{-i}$ is $u$ with the $i$th unit removed, $j \sim i$ if the units are defined as adjacent, $n_{\delta i} = |\{j:j \sim i\}|$ is the total number of adjacent units, and $\sigma > 0$ is the marginal standard deviation.
We follow recommendations of @freni2018note on scaling of precision matrices, disconnected adjacency graph components, and islands.
For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component $v^\star$ with proportion $\phi \in (0, 1)$ and spatially unstructured IID component $w^\star$ with proportion $1 - \phi$, both scaled to have generalised variance equal to one, and $\sigma > 0$ is the marginal standard deviation such that
$$
u = \sigma \left( \sqrt\phi \cdot v^\star + \sqrt{1 - \phi} \cdot w^\star \right).
$$

@@

We assume the household survey was run according to a complex survey design where each individual $j \in U$ has non-zero probability $\pi_j \in (0, 1)$ of appearing in the sample $S \subseteq U$. Suppose we observe an outcome $y_j$ for $j \in S$.
Let $w_j = 1 / \pi_j \times 1 / \omega_j$ be design weights, where $\omega_j$ is a non-response factor, then the weighted mean
$$
\hat y = \frac{\sum_{j \in S} w_j y_j}{\sum_{j \in S} w_j}.
$$
The Kish effective sample size is
$$
m = \frac{\left(\sum_{j \in S} w_j\right)^2}{\sum_{j \in S} w_j^2}.
$$
We make these computations using the `survey` \textsc{R} package [@lumley2004analysis].

@@

```{=tex}
\begin{table}[h]
\centering
\begin{tabularx}{\textwidth}{llXX}
\toprule
& Model section & Latent field & Hyperparameter \\ 
\midrule
\ref{sec:hiv-prev} & HIV prevalence & $36 + 3n$ & 9 \\ 
\ref{sec:art-cov} & ART coverage & $36 + 3n$ & 9 \\ 
\ref{sec:hiv-inc} & HIV incidence rate & $2 + n$ & 3 \\ 
\ref{sec:anc-test} & ANC testing & $4 + 2n$ & 2 \\ 
\ref{sec:art-attend} & ART attendance & $n$ & 1 \\ 
& Total & $78 + 7n$ & 24 \\ 
\bottomrule
\end{tabularx}
\label{tab:latent-hyper-size}
\caption{The numer of latent field parameters and hyperparameters in each model section.}
\end{table}
```

In this section, we describe the hyperparameter and latent field process specification for the model.
Whereas in the main text process and likelihood specifications are written together, here we consider the likelihood equations separately in Section \ref{sec:likelihood}.
Table \ref{tab:latent-hyper-size} gives the number of latent field parameters and hyperparameters in each section of the model.
Given a country with $n$ districts, then the total number of unique district-age-sex combinations is $n \times 2 \times 17 = 34n$.
In the case of Malawi then $n = 32$ such that the total number of latent field parameters is $491$ and the total number of hyperparameters is $24$.

@@

We model HIV prevalence $\rho_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
$$
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
$$
where $\beta^\rho_0 \sim \mathcal{N}(0, 5)$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}} \sim \text{AR}1(\sigma_{AS}^\rho, \phi_{AS}^\rho)$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(\sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
We use the prior distributions $\mathcal{N}^{+}(0, 2.5)$ for the standard deviation terms $\{\sigma_A^\rho, \sigma_{AS}^\rho, \sigma_X^\rho, \sigma_{XS}^\rho, \sigma_{XA}^\rho\}$, $\mathcal{U}(-1, 1)$ for the AR1 correlation parameters $\{\phi_A^\rho, \phi_{AS}^\rho\}$, and $\text{Beta}(0.5, 0.5)$ for the BYM2 proportion parameters $\{\phi_X^\rho, \phi_{XS}^\rho\}$.

<!-- Latent field size -->

<!-- 1 + 1 + 17 + 17 + n + n + n = 36 + 3n -->

<!-- Hyperparameter size -->

<!-- 5 + 2 + 2 -->

@@

We model ART coverage $\alpha_{x, s, a} \in [0, 1]$ on the logit scale using the linear predictor
$$
\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + u^\alpha_a + u_a^{\alpha, s = \text{M}} + u^\alpha_x + u_x^{\alpha, s = \text{M}} + u_x^{\alpha, a < 15} + \eta^\alpha_{R_x, s, a} 
$$
with terms and priors analogous to the HIV prevalence process model in Section \ref{sec:hiv-prev} above.

<!-- Latent field size -->

<!-- As above -->

<!-- Hyperparameter size -->

<!-- As above -->

### HIV incidence rate \label{sec:hiv-inc}

We model HIV incidence rate $\lambda_{x, s, a} > 0$ on the log scale using the linear predictor
$$
\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda, \label{eq:hiv-inc}
$$
where $\beta^\lambda_0 \sim \mathcal{N}(0, 5)$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}} \sim \mathcal{N}(0, 5)$ is the log incidence rate ratio for men compared to women,
$$
\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}}, \label{eq:rho-agg}
$$
is the HIV prevalence among adults 15-49 calculated by aggregating age-specific HIV prevalences, and
$$
\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}, \label{eq:alpha-agg}
$$
is the ART coverage among adults 15-49 calculated by aggregating age-specific ART coverages.
The term $\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ based on inputs to the Estimation and Projection Package (EPP) model [@eaton2019estimation], $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify fixed log incidence rate ratios by sex and age group calculated from Spectrum model output.

We model the proportion recently infected among HIV positive persons $\kappa_{x, s, a} \in [0, 1]$ as
$$
\kappa_{x, s, a} = 1 - \exp \left(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T \right),
$$
where $\Omega_T \sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T})$ is the mean duration of recent infection (MDRI), and $\beta_T \sim \mathcal{N}^{+}(\beta_{T_0}, \sigma^{\beta_T})$ is the false recent ratio (FRR).
We use an informative prior on $\Omega_T$ based on the characteristics of the recent infection testing algorithm (RITA).
For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$. 
For PHIA surveys we assume there is no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.

<!-- Latent field size -->

<!-- 1 + 1 + n + 1 + 1 = 4 + n -->

<!-- Hyperparameter size -->

<!-- 3 -->

### ANC testing \label{sec:anc-test}

HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women are modelled as being offset on the logit scale from the corresponding district-age indicators $\rho_{x, F, a}$ and $\alpha_{x, F, a}$ according to
\begin{align*}
\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}}, \\
\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}},
\end{align*}
where, for $\theta \in \{\rho, \alpha\}$, $\beta^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, 5)$ are the average differences between population and ANC outcomes, $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ are IID district random effects with $\sigma_X^{\theta^{\text{ANC}}} \sim \mathcal{N}^+(0, 1)$, and $\eta_{R_x, a}^{\theta^{\text{ANC}}}$ for are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$.

For adult women 15-49 we model the number of ANC clients $\Psi_{x, a} > 0$ on the log scale
$$
\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi
$$
where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are fixed age-sex fertility ratios in Spectrum region $R_x$, $\beta^\psi$ are fixed log rate ratios for the number of ANC clients relative to the predicted fertility, and $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects, with $\sigma^\psi \sim \mathcal{N}^+(0, 1)$.

<!-- Latent field size -->
<!-- 1 + n + 1 + n = 2 + 2n -->
<!-- Hyperparameter size -->
<!-- 2 -->

### ART attendance \label{sec:art-attend}

@@

Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person on ART residing in district $x$ receives ART in district $x'$.
We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district or its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation, and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.
To model $\gamma_{x, x'}$ for $x \sim x'$ we use a multinomial logistic regression model, based on the log-odds ratios
\begin{equation}
\tilde \gamma_{x, x'} = \log \left( \frac{\gamma_{x, x'}}{1 - \gamma_{x, x'}} \right) = \tilde \gamma_0 + u_x^{\tilde \gamma}, \label{eq:log-or}
\end{equation}
where $\tilde \gamma_0 = -4$ is a fixed intercept, and $u_x^{\tilde \gamma} \sim \mathcal{N}(0, \sigma_X^{\tilde \gamma})$ are district random effects with $\sigma_X^{\tilde \gamma} \sim \mathcal{N}^+(0, 2.5)$.
Note that Equation \ref{eq:log-or} does not depend on $x'$, such that $\gamma_{x, x'}$ is only a function of $x$.
Choice of $\tilde \gamma_0 = -4$ implies a prior mean on $\gamma_{x, x'}$ of 1.8%, such that $(100 - 1.8 \times \text{ne}(x))\%$ of ART clients in district $x$ obtain treatment in their home district, a-priori.
We fix $\tilde \gamma_{x, x} = 0$ and recover the multinomial probabilities using the softmax
$$
\gamma_{x, x'} = \frac{\exp(\tilde \gamma_{x, x'})}{\sum_{x^\star \in \{x, \text{ne}(x)\}} \exp(\tilde \gamma_{x, x^\star})}.
$$
Given the total number of PLHIV on ART $A_{x, s, a} = N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}$, the number of ART clients who reside in district $x$ and obtain ART in district $x'$ are $A_{x, x', s, a} = A_{x, s, a} \cdot \gamma_{x, x'}$, and the total attending ART facilities in district $x'$ are 
$$
\tilde A_{x', s, a} = \sum_{x \in \{x', \text{ne}(x')\}} A_{x, x', s, a}.
$$

## Likelihood specification \label{sec:likelihood}

@@

For HIV prevalence, ART coverage and recent HIV infections, denoted by $\theta \in \{\rho, \alpha, \kappa\}$, the household survey $\nu$ furnishes weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}}$ with respective Kish effective sample sizes $m^{\hat \theta}_{\{x\}, \{s\}, \{a\}}$, and observed number of cases $y^{\hat \theta}_{\{x\}, \{s\}, \{a\}} = m^{\hat \theta}_{\{x\}, \{s\}, \{a\}} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}}$.
We use following three binomial working likelihoods
\begin{align*}
y^{\hat \rho}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \rho}_{\{x\}, \{s\}, \{a\}}, \rho_{\{x\}, \{s\}, \{a\}}), \quad
\rho_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a}}, \\
y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \alpha}_{\{x\}, \{s\}, \{a\}}, \alpha_{\{x\}, \{s\}, \{a\}}), \quad
\alpha_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}. \\
y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}} &\sim \text{xBin}(m^{\hat \kappa}_{\{x\}, \{s\}, \{a\}}, \kappa_{\{x\}, \{s\}, \{a\}}), \quad
\kappa_{\{x\}, \{s\}, \{a\}} = \frac{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \kappa_{x, s, a}}{\sum_{\{x\}} \sum_{\{s\}} \sum_{\{a\}} N_{x, s, a} \cdot \rho_{x, s, a}},
\end{align*}
The generalised binomial $y \sim \text{xBin}(m, p)$ is defined for $y, m \in \mathbb{R}^+$ with $y \leq m$ such that
$$
\log p(y) = \log \Gamma(m + 1) - \log \Gamma(y + 1) - \log \Gamma(m - y + 1) + y \log p + (m - y) \log(1 - p),
$$
where the gamma function $\Gamma$ is such that $\forall n \in \mathbb{N}$, $\Gamma(n) = (n - 1)!$.
